# Course overview {.unnumbered}

```{r, echo=FALSE}
library(kableExtra)
```


```{r, echo=FALSE}
session_data <- cbind.data.frame(
  Date = as.Date(
    c("2024.10.18", "2024.10.25", 
           "2024.11.08", "2024.11.15", "2024.11.22", "2024.11.29",
           "2024.12.06", "2024.12.13", "2024.12.20",
           "2025.01.10", "2025.01.17"),
    "%Y.%m.%d"),
  Block = c(rep("Getting started", 2), rep("Programming in R", 2), 
            rep("Data Science in R", 3), rep("Databases", 2), rep("GIS", 2)),
  Topic = c("Data and research data management",
            "Git and GitHub",
            "Introduction to R",
            "Best practices in programming",
            "Data Science Workflow",
            "Count data and seriation",
            "Compositional data",
            "Databases (I)",
            "Databases (II)",
            "GIS (I)",
            "GIS (II)")
)
session_data$Block <- factor(session_data$Block, unique(session_data$Block))
```

040468 Datenmanagement und digitale Archäologie (ÜB, unterrichtet in Englisch) Übungen (3 CP)

**Time slot** / **_Zeitfenster_**: Fr 14-16 Uhr c.t.

**Place** / **_Ort_**: Raum 2

**Course instructors** / **_Kursleiter_**: Andreas Angourakis / Tim Klingenberg

**Support** / **_Unterstützung_**: Thomas Rose

## **Course schedule**/**_Kursplan_**:

```{r, echo=FALSE}
schedule_table <- knitr::kable(session_data[,c("Date", "Topic")], row.names = TRUE)
schedule_table <- kableExtra::pack_rows(schedule_table, index = table(session_data$Block), background = "#a9a9a9")
schedule_table
```

## **Evaluation** / **_Kursbewertung_**

(Attendance and final examination)

## Acknowledgements

The conception of the course structure, as well as the short summaries, exercises, and images shown in each chapter, greatly benefited from [Large Language Models](https://en.wikipedia.org/wiki/Large_language_model) used as companion writer and programmer. As such, we own greatly to the current richness of reference information freely available on Internet.

The models and services used are:

- [ChatGPT](https://chatgpt.com/) (GPT-4o) by OpenAI for brainstorming, text and code writing suggestions, collection and articulation of references.
- [WebChatGPT](https://tools.zmo.ai/webchatgpt), a free browser extension that enhances ChatGPT by providing Internet access directly within the chat interface, used to aid Internet search.
- [Leonardo.ai](https://app.leonardo.ai/) (user tokens) for generating purely aesthetic visual assets.
